\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-17 chapters/bttda.tex
Indicated comments Reinhold and Fabien

Recent research has shown that supervised decoders adopting a more flexible structure
can improve \ac{bci} performance.
Promising results have been achieved for regression tasks using
Higher Order Partial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022a}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaves room for improvement.
Instead of optimizing features directly for class separability, a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in these regression approaches are still more constrained
than what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low-rank common subspace
between the input and classification labels.
\textcite{Huang2020} propose a supervised approach for finding multiple discriminant
multilinear spectral filter terms and apply it to motor imagery BCI, but their
decomposition is also restricted in flexibility.

@@

The \ac{hoda} projection in \cref{eq:HODA-backward} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model, visualized in \cref{fig:HODA-forward}, is given by:
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$, and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[A forward projection for \ac{hoda}.]{The forward projection for HODA.
    By calculating activation patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\text{E}\left[
      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
    \right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \hat{\ten{X}}(n)\right\]^2
  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right\]^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \cref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\cite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
  \centering
	\footnotesize
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \cite{Chevallier2024}. Accuracy
    is listed for BNCI2014-001, and area under the receiver-operator
    characteristic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
    for BNCI2014-001 but does not reach performance comparable to the current
    state-of-the-art. For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA.

\subsection{Block contribution}
To analyze the contribution of extra feature blocks extracted by BTTDA over
the first one found by HODA, we pick an ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation, and test data as a function
of the number of blocks increases, shown in \cref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\hat{\ten{X}}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as:
\begin{equation}
  \nmse\left(\ten{X}, \hat{\ten{X}}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
  \centering
  \makebox[\linewidth][c]{
	  \input{figures/bttda/blocks.pgf}
  }
  \caption[Analysis of NMSE and classification score per block.]{%
    Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}

@@

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or higher than HODA+LDA.
The classification score for BTTDA+LDA was also slightly higher than that of
restricted PARAFACDA+LDA for the ERP datasets and one of the MI datasets, but these
results were not significant, and further studies with more datasets and
subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

@@

Via an accompanying heuristic model selection procedure, \ac{bci} decoders using
\ac{bttda} feature extraction can significantly outperform decoders based on
\ac{hoda} and reach state-of-the-art decoding performance on event-related potential
problems (second-order tensors) and scores on par with or higher than \ac{hoda} in motor
imagery problems (third-order tensors).
Moving from the rigid Tucker tensor structure of \ac{hoda} to the more flexible
block-term structure shifts the problem from finding optimally constrained
multilinear projections to model and feature selection.

>>>>

Recent research has shown that supervised decoders adopting a more flexible structure
can improve \ac{bci} performance.
Promising results have been achieved for regression tasks using
Higher Order Partial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022a}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaves room for improvement.
Instead of optimizing features directly for class separability, a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in these regression approaches are still more constrained
than what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low-rank common subspace
between the input and classification labels.
\textcite{Huang2020} propose a supervised approach for finding multiple discriminant
multilinear spectral filter terms and apply it to motor imagery BCI, but their
decomposition is also restricted in flexibility, since the solution can only
extract rank $(r_1,r_2,1)$ with mode 3 corresponding to the frequency domain.

@@

The \ac{hoda} projection in \cref{eq:HODA-backward} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model, visualized in \cref{fig:HODA-forward}, is given by:
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$, and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[A forward projection for \acs{hoda}.]{The forward projection for HODA.
    By calculating activation patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\text{E}\left[
      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
    \right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \hat{\ten{X}}(n)\right]^2
  = \argmin_{\{\mat{A}\}}\sum_n^N\left[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right]^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \cref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\cite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
  \centering
	\footnotesize
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \cite{Chevallier2024}. Accuracy
    is listed for BNCI2014-001, and area under the receiver-operator
    characteristic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
    for BNCI2014-001 but does not reach performance comparable to the current
    state-of-the-art. For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA\todo{qualitative}.

\subsection{Block contribution}
To analyze the contribution of extra feature blocks extracted by BTTDA over
the first one found by HODA, we pick an ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation, and test data as a function
of the number of blocks, shown in \cref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\hat{\ten{X}}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as:
\begin{equation}
  \nmse\left(\ten{X}, \hat{\ten{X}}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
  \centering
  \makebox[\linewidth][c]{
	  \input{figures/bttda/blocks.pgf}
  }
  \caption[Analysis of NMSE and classification score per block.]{%
    Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}

@@

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or
higher than HODA+LDA\todo{qualitative}.
The classification score for BTTDA+LDA was also slightly
higher\todo{qualitative} than that of
restricted PARAFACDA+LDA for the ERP datasets and one of the MI datasets, but these
results were not significant, and further studies with more datasets and
subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

@@

Via an accompanying heuristic model selection procedure, \ac{bci} decoders using
\ac{bttda} feature extraction can significantly outperform decoders based on
\ac{hoda} and reach state-of-the-art decoding performance on event-related potential
problems (second-order tensors) and scores on par with or higher than \ac{hoda} in motor
imagery problems (third-order tensors)\todo{qualitative}.
Moving from the rigid Tucker tensor structure of \ac{hoda} to the more flexible
block-term structure shifts the problem from finding optimally constrained
multilinear projections to model and feature selection.

### 2024-11-17 chapters/conclusion.tex
Indicated comments Reinhold and Fabien

\subsection{Developed \acs{erp} decoders}
\label{sec:conclusion/contrib/decoders}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\textcite{VanDenKerchove2022}.

@@

Finally, we developed the \acf{wcble} decoder, specifically designed to address
the challenges posed by P3 latency jitter in covert \ac{vsa} settings.
Latency variability is a well-known issue in \ac{erp} decoding, particularly
when users engage in covert attention, where the timing of the P3 component
fluctuates significantly across trials.
Traditional decoders struggle in these settings due to the inconsistent timing
of brain responses, which can reduce the robustness and accuracy of
\ac{erp} decoding.
The \ac{wcble} decoder mitigates this issue by introducing an iterative process
to estimate and align \ac{erp} latencies across trials.
This is performed in such a way that discriminative power of the \ac{erp}
signal is preserved and enhanced.
The method has been published in~\textcite{VanDenKerchove2024}.
We designed this method to improve decoding where
unseen incoming test data is not yet known.
\Ac{wcble} was first tested on synthetic data to evaluate its
effectiveness under conditions of controlled latency variability and noise.
These results showed that it is robust to higher noise and jitter compared to
a non-iterative method in a latency estimation task.
It was also shown that decoding accuracy was higher in high jitter and noise
settings compared to the non-iterative method and a state-of-the-art decoder.

@@

We evaluated our proposed \ac{wcble} algorithm, its non-iterative
counterpart, and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
We evaluated the BCI decoding performance in a single-trial classification experiment,
as well as in a target selection experiment reflecting BCI operation.
Performance was significantly different between decoders, but this
result was significantly dependent on the \ac{vsa} condition and the dataset.
While \ac{wcble} was outperformed by the state-of-the-art in overt \ac{vsa}
decoding, it increased covert \ac{vsa} decoding.

@@

To further study the gaze-independent performance of these algorithms, transfer
learning between \ac{vsa} conditions was studied to simulate conditions where an
individual with gaze impairment can end up in different \ac{vsa} settings within
a \ac{bci} operation session due to their lack of proper motor control.
When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
results in a significant decrease in performance compared to the state-of-the-art,
consistent with the within-condition results.
For all other pairs of training and evaluation VSA conditions, however,
\ac{wcble} was equal to or significantly better.

@@

Finally, we wish to expand our experimentation with individuals with
\acp{sspgi}.
We aim to revisit the participants with an on-line experiment and comfort
assessment to establish the usability of the proposed interface and compare its
\ac{itr} to the state-of-the-art literature.

@@

Throughout this thesis, we've noticed that linear or multilinear decoding
methods consistently outperform non-linear methods, such as SVMs or those based
on Riemannian Geometry.
Within these linear models, regularization (shrinkage, covariance structure,
tensor structure/rank) is of paramount importance, since neural data is
inherently multivariate and \ac{bci} calibration time is minimized.
Proper regularization should impose some specific structure on the linear model,
ideally reflecting properties of the signal.
When faced with a new problem and few data relative to dimensionality, it is a
good first instinct to pick a simple, restrictive structure.

@@

Unfortunately, this is also probably one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
This approach can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a working, long-term
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and doesn’t work for them, and where research efforts
should be focused.

Similarly, there is value in maintaining a working, on-line in-house \ac{bci}
system that can be iterated upon and adapted if necessary.
Experiments can be supported by advance calculations of hypothetical
\ac{itr}, ensuring that on-line or off-line experiments reasonably reflect
real-world operation, and no effort is wasted exploring settings that will not
achieve satisfactory performance.
For example, why study a specific number of repetitions or \ac{isi}, if they
would not be practical to achieve a high \ac{itr}, but could still influence
decoding performance?

This top-down approach is probably not suited for all types of \ac{bci}
research.
Yet, in visual oddball \ac{bci}, where the challenge no longer lies in generating
a proof of concept but in translating the concept into useful technology, this
approach might help avoid prematurely solving problems that may not necessarily
arise further down the road.
A top-down approach could improve the quality of study outcomes by enabling the
reporting of realistic, interpretable metrics,
and by taking into account the human aspect of working \emph{with} \ac{bci} users.

>>>>

\subsection{Developed \acs{erp} decoders}
\label{sec:conclusion/contrib/decoders}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher
accuracy\todo{qualitative} in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\textcite{VanDenKerchove2022}.

@@

Finally, we developed the \acf{wcble} decoder, specifically designed to address
the challenges posed by P3 latency jitter in covert \ac{vsa} settings.
Latency variability is a well-known issue in \ac{erp} decoding, particularly
when users engage in covert attention, where the timing of the P3 component
fluctuates significantly across trials.
Traditional decoders struggle in these settings due to the inconsistent timing
of brain responses, which can reduce the robustness and accuracy of
\ac{erp} decoding.
The \ac{wcble} decoder mitigates this issue by introducing an iterative process
to estimate and align \ac{erp} latencies across trials.
This is performed in such a way that discriminative power of the \ac{erp}
signal is preserved and enhanced.
The method has been published in~\textcite{VanDenKerchove2024}.
We designed this method to improve decoding where
unseen incoming test data is not yet known.
\Ac{wcble} was first tested on synthetic data to evaluate its
effectiveness under conditions of controlled latency variability and noise.
These results showed that it is robust to higher noise and jitter\ compared to
a non-iterative method in a latency estimation task\todo{qualitative}.
Decoding accuracy was higher in high jitter and noise
settings compared to the non-iterative method and a state-of-the-art
decoder\todo{qualitative}.

@@

We evaluated our proposed \ac{wcble} algorithm, its non-iterative
counterpart, and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
We evaluated the BCI decoding performance in a single-trial classification experiment,
as well as in a target selection experiment reflecting BCI operation.
Performance was significantly different between decoders, but this
result was significantly dependent on the \ac{vsa} condition and the dataset.
While \ac{wcble} was outperformed by the state-of-the-art in overt \ac{vsa}
decoding, it increased covert \ac{vsa} decoding.\todo{qualitative}

@@

To further study the gaze-independent performance of these algorithms, transfer
learning between \ac{vsa} conditions was studied to simulate conditions where an
individual with gaze impairment can end up in different \ac{vsa} settings within
a \ac{bci} operation session due to their lack of proper motor control.
When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
results in a significant decrease in performance compared to the state-of-the-art,
consistent with the within-condition results.\todo{qualitative}
For all other pairs of training and evaluation VSA conditions, however,
\ac{wcble} was equal to or significantly better.\todo{qualitative}

@@

Finally, we wish to expand our experimentation with individuals with
\ac{sspgi}.
We aim to revisit the participants with an on-line experiment and comfort
assessment to establish the usability of the proposed interface and compare its
\ac{itr} to the state-of-the-art literature.

@@

Throughout this thesis, we noticed that linear or multilinear decoding
methods consistently outperform non-linear methods, such as SVMs or those based
on Riemannian Geometry.
Within these linear models, regularization (shrinkage, covariance structure,
tensor structure/rank) is of paramount importance, since neural data is
inherently multivariate and \ac{bci} calibration time is minimized.
Proper regularization should impose some specific structure on the linear model,
ideally reflecting properties of the signal.
When faced with a new problem and few data relative to dimensionality, it is a
good first instinct to pick a simple, restrictive structure.

@@

Unfortunately, this is also probably one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
This approach can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a long-term working
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and doesn’t work for them, and where research efforts
should be focused.

Similarly, there is value in maintaining a working, on-line, in-house \ac{bci}
system that can be iterated upon and adapted if necessary.
Given this system's design and data gathered from earlier use, experiments can
be supported by advance calculations of hypothetical
\ac{itr}, ensuring that on-line or off-line experiments reasonably reflect
real-world operation, and no effort is wasted exploring settings that will not
achieve satisfactory performance.
For example, why study a specific number of repetitions or \ac{isi}, if they
would not be practical to achieve a high \ac{itr}, but could still influence
decoding performance?

This top-down approach is probably not suited for all types of \ac{bci}
research.
Yet, in visual oddball \ac{bci}, where the challenge no longer lies in generating
a proof of concept but in translating the concept into useful technology, this
approach might help avoid prematurely solving problems that may not necessarily
arise further down the road.
A top-down approach could improve the quality of study outcomes by enabling the
reporting of realistic, interpretable metrics,
and by taking into account the human aspect of working \emph{with} instead of
\emph{for} \ac{bci} users.
